---
title: "Clustering"
author: "Thomas Amble Karlsen and Eirik Sørhus"
date: "`r Sys.Date()`"
output: 
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, 
                      warning = F, 
                      message = F)
```

```{r tema, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Laster ned pakken med formatet for dokumentet i tilfelle pakken ikke allerede er installert
```

Libraries
```{r }
library(tidyverse)
library(usedist)
library(dendextend)
```

## Loading the data
```{r }
load("Data/AspWood_normalized.RData")
```

Identifies genes with expression over 5 and under 1 to be regulated genes. Filtering so vst contains only the "regulated" genes and stored in a new matrix aspwood. (Finding the data which varies between samples)
```{r }
vst %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Gene") %>%          # adding rownames
  pivot_longer(-Gene, names_to = "Sample", values_to = "Expression") %>% 
  mutate(Expressed = Expression > 5,
         Silent = Expression < 1) %>%
  group_by(Gene) %>% 
  summarise(Expressed = sum(Expressed),
            Silent = sum(Silent)) %>% 
  filter(Expressed >= 1, Silent >= 1) %>% 
  pull(Gene) -> regulated.genes

aspwood <- vst[rownames(vst) %in% regulated.genes,]
```


Plotting the expression profile on a random gene
```{r }
set.seed(3)

aspwood %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Gene") %>%          # Adding rownames
  filter(Gene == sample(regulated.genes, 1)) %>% 
  pivot_longer(-Gene, names_to = "Sample", values_to = "Expression") %>% 
  separate(Sample, into = c("Tree", "Sample"), sep = "-") %>% 
  mutate(Sample = as.numeric(Sample)) %>% 
  ggplot(aes(x = Sample, y = Expression)) +
  geom_line(linewidth = 1.5) +
  theme_bw()
```


## Hierarchical clustering
Adding stages of growth/life to the aspwood data
```{r }
aspwood.stages <- factor(c(rep("PhloemZone", 5), rep("ExpansionZone", 5), rep("SecondaryCellWallZone", 9), rep("LignificationZone", 6)))
```

Preforming hierarchical clustering
```{r }
aspwood.tree <- aspwood %>% 
  t() %>% #clustering for samples not genes
  dist() %>% # Eucleadian distance
  hclust() %>% # preforming the clustering
  as.dendrogram() %>% # making it a denogram
  reorder(1:25, agglo.FUN = mean) %>% #reorganises the denogram
  as.hclust() # converting it back to hierarchical clustering

# plotting
aspwood.tree %>% 
  as.dendrogram() %>% 
  set("labels_cex", 1) %>% # label size
  set("leaves_pch", 19) %>%  # node point type
  set("leaves_cex", 0.75) %>%  # node point size
  set("leaves_col", aspwood.stages[aspwood.tree$order]) %>%  # node point color
  set("branches_k_color", k=4) %>% # branch color
  plot()
```

## Heatmap

```{r }
ADF <- data.frame(row.names = colnames(aspwood), # Removing row names
                  Stages = aspwood.stages)

heatmap(aspwood,
         scale = "row", # Scaling rows, not columns
         cluster_cols = aspwood.tree,
         cluster_rows = T,
         clustering_distance_rows = "correlation", 
         clustering_method = "ward.D", # Using Ward's method, as they did in the paper
         annotation_col = ADF,
         show_rownames = F)
```

## K-means clustering
```{r }
aspwood.kmeans <- kmeans(t(aspwood), 4)
table(aspwood.stages, aspwood.kmeans$cluster)
```
The k-means clustering method is not clustering as well as the hierarchial clustering method. This can be seen in the table above, as there is 8 samples in the cluster for SecondaryCellWallZone, instead of 9, which it is in reality. The remaining sample from this cluster has instead ended up in the cluster of samples from the ExpansionZone.


## Principle component analysis (PCA)
```{r }
aspwood.pca <- prcomp(t(aspwood))
```

```{r }
aspwood.pca$x %>% 
  ggplot(aes(PC1,
             PC2,
             color = aspwood.stages)) +
  geom_point()
```

shows a summary of the principle components, shows SD, proportion of variance and cumulative proportion
```{r }
summary(aspwood.pca)
```
