---
title: "Transcriptomics"
author: "Thomas Amble Karlsen and Eirik Sørhus"
date: "`r Sys.Date()`"
output: 
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    code_folding: hide
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, 
                      warning = F, 
                      message = F)
```

```{r tema, include=FALSE}
if(!require(rmdformats)) install.packages("rmdformats",repos = "http://cran.us.r-project.org")
# Laster ned pakken med formatet for dokumentet i tilfelle pakken ikke allerede er installert
```

## Loading Data
```{r Reading file}
library(tidyverse)
```

```{r}
samples <- list.files("Data/AspWood/AspWood")

counts <- tibble()
for (sample in samples) {
  file <- paste0("Data/AspWood/AspWood/", sample, "/quant.sf")
  
  sample.trimmed <- gsub("_sortmerna_trimmomatic", "", sample)
  
  c <- read_tsv(file) %>%
    select(Name, NumReads) %>%
    rename(Genes = Name, !!sym(sample.trimmed) := NumReads)
  
  if (sample == samples[1]) {
    counts <- c 
  } else {
    counts <- cbind(counts, c %>% select(-Genes))
  }
}

```


```{r Finding Counts}
# Method 1
counts <- counts %>% 
  pivot_longer(cols = -Genes, names_to = "Samples", values_to = "Expression") %>%
  separate(Genes, into = c("Genes"), sep = "\\.", extra = "drop") %>% 
  group_by(Genes, Samples) %>%
  summarise(Expression = sum(Expression)) %>%
  pivot_wider(names_from = "Samples", values_from = "Expression")

# Method 2
counts <- counts %>% 
  separate(Genes, into = c("Genes"), sep = "\\.", extra = "drop") %>% 
  group_by(Genes) %>%
  summarise_if(is.numeric, sum)
```

```{r reshaping}
counts.long <- counts %>% 
  pivot_longer(cols = -Genes, names_to = "Samples", values_to = "Expression")
```

## Exploring teh count data and log-transform
Dotplot of counts. The x-axis is the samples and y-axis is expression levels. The dots represent counts.
```{r dotplot not log or normelized}
ggplot(counts.long, 
       aes(x = Samples,
           y = Expression)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90))
```

A boxplot with expression as y-axis and samples as x-axis.
```{r box plot not log og normelized}
ggplot(counts.long, 
       aes(x = Samples,
           y = Expression)) +
  geom_boxplot(outliers = F) +
  theme(axis.text.x = element_text(angle = 90))
```

Boxplot, same as over, but the expression values are log2-transformed. log2(counts + 1) is used due to many values are zero.
```{r box plot log transformed not normelized}
ggplot(counts.long, 
         aes(x = Samples,
             y = log2(Expression+1))) +
  geom_boxplot(outliers = F) +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(y = "Expression")
```

## Normalization
Makes a DEseq2 dataset and normelizes it
```{r}
library(DESeq2)

counts.mat <- counts %>% 
  column_to_rownames(var = "Genes") %>% 
  as.matrix() %>% 
  round()

dds <- DESeq2::DESeqDataSetFromMatrix(countData = counts.mat,
                              colData = data.frame(conditions = as.factor(colnames(counts.mat))),
                              design = ~ conditions)

dds <- DESeq2::estimateSizeFactors(dds)
```

Boxplot with normalized data and log2 transformed
```{r box plot normalized and log}
DESeq2::counts(dds, normalized=TRUE) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Genes") %>%  
  pivot_longer(cols = -Genes, names_to = "Samples", values_to = "Expression") %>% 
  mutate(Expression = log2(Expression + 1)) %>% 
  ggplot(aes(x = Samples, y = Expression)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Varianze stabelizing transformation (VST)
Varianze stabelizing transforming the data and making the min value 0
```{r}
vst <- DESeq2::varianceStabilizingTransformation(counts.mat)
vst <- vst - min(vst)
```

boksplot with VST normalized date and log2 transformed expression
```{r}
vst %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Genes") %>%  
  pivot_longer(cols = -Genes, names_to = "Samples", values_to = "Expression") %>% 
  mutate(Expression = log2(Expression + 1)^2) %>% # Squaring the log-transformed transformation to make it visually comparable to previous boxplots
  ggplot(aes(x = Samples, y = Expression)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Removing lowly expressed genes
removing genes with expression less than 10
```{r}
counts.long %>% 
  group_by(Genes) %>% 
  summarise(Expression = sum(Expression)) %>% 
  filter(Expression < 10) -> lowgenes


# Removing genes with less than 10 mapped reads:
vst <- vst[!(rownames(vst) %in% lowgenes),]
```

## Looking at data
The plots below are the expression of the genes over time (samples are from different times)
```{r gene SUS6: Potra2n4c9149}
vst %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Genes") %>%  
  pivot_longer(cols = -Genes, names_to = "Samples", values_to = "Expression") %>% 
  mutate(Expression = log2(Expression + 1)) %>% 
  filter(Genes == "Potra2n4c9149") %>% 
  ggplot(aes(x = Samples,
             y = Expression,
             group = Genes)) +
  geom_line(linewidth = 2) +
  labs(y = "Expression (VST)",
       title = "SUS6: Potra2n4c9149") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust=1),
        panel.border = element_rect(fill=NA))
```

```{r gene CDC2: Potra2n16c30563}
vst %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "Genes") %>%  
  pivot_longer(cols = -Genes, names_to = "Samples", values_to = "Expression") %>% 
  mutate(Expression = log2(Expression + 1)) %>% 
  filter(Genes == "Potra2n16c30563") %>% 
  ggplot(aes(x = Samples,
             y = Expression,
             group = Genes)) +
  geom_line(linewidth = 2) +
  labs(y = "Expression (VST)",
       title = "CDC2: Potra2n16c30563") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, 
                                   vjust = 0.5, 
                                   hjust=1),
        panel.border = element_rect(fill=NA))
```

